\LaTeXe\ guide for the Journal of Functional Programming

### 2019-10-19 papers/jfp/3-abstractions.tex
Polish on S3-4

\noindent
We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.) The
task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values. It pattern-matches on the key to see if it has a task
description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating that the key is an input. If there is a formula in the
cell, it returns the \hs{Task} to compute the value of the formula. Every
task is given a \emph{callback} \hs{fetch} to find the value of any keys on
which it depends. To run a \hs{Task}, we simply apply the function it holds to a
callback (see the definition of the function \hs{run} in Fig.~\ref{fig-types}).

@@

\noindent
The \hs{busy} build system defines the callback \hs{fetch} so that, when given a
key, it brings the key up to date in the store, and returns its value.
The function \hs{fetch} runs in the standard Haskell \hs{State} monad -- see
Fig.~\ref{fig-stdlib} -- initialised with the incoming \hs{store} by \hs{execState}.
To bring a key up to date, \hs{fetch} asks the task description \hs{tasks} how
to compute the value of \hs{k}. If \hs{tasks} returns \hs{Nothing} the key is an
input, so \hs{fetch} simply reads the result from the store. Otherwise \hs{fetch}
runs the obtained \hs{task} to produce a resulting value~\hs{v}, records the new
key/value mapping in the store, and returns \hs{v}. Notice that \hs{fetch}
passes itself to \hs{task} as an argument, so that the latter can use \hs{fetch}
to recursively find the values of \hs{k}'s dependencies.

@@

\subsection{The Need for Polymorphism in Task}\label{sec-why-polymorphism}
% \vspace{-0.5mm}

The previous example illustrates why the \hs{Task} abstraction is polymorphic
in~\hs{f}; recall its definition from~\S\ref{sec-task}:

\begin{minted}[xleftmargin=5pt]{haskell}
newtype Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
\end{minted}

@@

But the task description cannot possibly work for \emph{any} \hs{f} whatsoever; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c} and, in exchange, the task has access to the operations of class \hs{c}.
So the type \hs{Task} emerges naturally, almost inevitably.
But now that it \emph{has} emerged, we find that constraints \hs{c} classify
task descriptions in a very interesting, and practically useful, way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
  \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: a functorial task
  description cannot even use the application operator \hs{<*>}, which limits
  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}
  (ignoring the recent multi-stage builds). It
  is interesting to note that, when run on such a task description, the
  \hs{busy} build system will build each key at most once, thus partially
  fulfilling the minimality requirement~\ref{def-minimal}. Alas, it still has no
  mechanism to decide which input keys changed since the previous build.
\item \hs{Task}~\hs{MonadFail} correspond to monadic tasks that may fail. For
  example, the formula \cmd{A1/A2} may fail due to division by zero. We will
  discuss this in~\S\ref{sec-failures}.
\item \hs{Task}~\hs{MonadPlus}, \hs{Task}~\hs{MonadRandon} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

@@

\noindent
The big difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
\hs{fetch} \emph{different keys} depending on whether or not \hs{c1}~\hs{==}~\hs{1}.

Since the \hs{busy} build system introduced in~\S\ref{sec-general-build} always
rebuilds every dependency it encounters, it is easy for it to handle dynamic
dependencies. For minimal build systems, however, dynamic dependencies, and hence
monadic tasks, are much more challenging, as we shall see
in~\S\ref{sec-implementations}.

@@

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
Here is our first use of polymorphism in \hs{f}, discussed in~\S\ref{sec-why-polymorphism}.
The use of \hs{Identity} just fixes the ``impedance mismatch'' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad and pass to the \hs{task}. The result has type \hs{Identity}~\hs{v}, which
we unwrap with \hs{runIdentity}.

@@

The presented definition of correctness needs to be adjusted for
non-deterministic tasks and shallow cloud builds, as will be discussed
in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects},
respectively.

@@

\noindent
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-stdlib}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-stdlib}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is what
\hs{dependencies} does:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
["A1","A2"]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B2"
["B1"]
\end{minted}
\vspace{1mm}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer

track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [(k, v)])
track task fetch = runWriterT $ run task trackingFetch
  where
    trackingFetch :: k -> WriterT [(k, v)] m v
    trackingFetch k = do v <- lift@\,@(fetch k); tell@\,@[(k, v)]; return@\,\blk{v}@
\end{minted}
\vspace{1mm}

\noindent
This function uses the standard Haskell \hs{WriterT} \emph{monad
transformer}~\cite{liang1995monad} for recording additional
information~--~a list of key/value pairs \hs{[(@@k,}~\hs{v)]}~--~when executing
a task in an arbitrary monad~\hs{m} (see the types in Fig.~\ref{fig-stdlib}).
We substitute the given \hs{fetch} with a
\hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key/value pair. The \hs{task} returns a value of type
\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap with
\hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
with dynamic dependencies, see~\S\ref{sec-implementation-shake}. Here we show an
example of \hs{track}ing monadic tasks when \hs{m}~\hs{=}~\hs{IO}, by defining
a corresponding \hs{fetchIO} of type \hs{String}~\hs{->}~\hs{IO}~\hs{Integer},
which allows us to demonstrate the dynamic nature of monadic dependencies in
GHCi.

@@

\noindent
Here the keys $n < 2$ are input parameters, and one can obtain the usual
Fibonacci sequence by picking $F_0=0$ and $F_1=1$, respectively. Any minimal
build system will compute the sequence with memoization, i.e. without
recomputing the same value twice.

@@

\noindent
The main case ($m>0 \wedge n>0$) starts by looking up the pair of indices
\hs{(}\hs{m,}~\hs{n)} in the \hs{cache}. If the cache has \hs{Nothing}, we
calculate the resulting \hs{value} as before and \hs{modify} the cache
accordingly; otherwise, if we have a cache hit, we return the obtained value
immediately, skipping the actual calculation and thus potentially saving a large
amount of work. Indeed you do not want to recompute $A(4,2)=2^{65536}-3$
unnecessarily; all of its 19,729 decimal digits have already been helpfully
computed, e.g. see~\cite{ackermann42}. We will use \hs{MonadState} tasks in our
models of cloud build systems in~\S\ref{sec-implementations}.

>>>>

\noindent
We instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.) The
task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values. It pattern-matches on the key to see if it has a task
description (in the \Excel case, a formula) for it. If not, it returns
\hs{Nothing}, indicating that the key is an input. If there is a formula in the
cell, it returns the \hs{Task} to compute the value of the formula. Every
task is given a \emph{callback} \hs{fetch} to find the value of any keys on
which it depends. To run a \hs{Task}, we simply apply the function it holds to a
suitable callback (see the definition of the function \hs{run} in
Fig.~\ref{fig-types}).

@@

\noindent
The \hs{busy} build system defines the callback \hs{fetch} that, when given a
key, brings the key up to date in the store, and returns its value.
The function \hs{fetch} runs in the standard~Haskell \hs{State} monad
(Fig.~\ref{fig-stdlib}) initialised with the incoming \hs{store} by \hs{execState}.
To bring a key~\hs{k} up to date, \hs{fetch} asks the task description \hs{tasks}
how to compute its value. If \hs{tasks} returns \hs{Nothing} the key is an
input, so \hs{fetch} simply reads the result from the store. Otherwise \hs{fetch}
runs the obtained \hs{task} to produce a resulting value~\hs{v}, records the new
key/value mapping in the store, and returns \hs{v}. Notice that \hs{fetch}
passes itself to \hs{task} as an argument, so the latter can use \hs{fetch}
to recursively find the values of \hs{k}'s dependencies.

@@

To avoid the recomputation, \hs{busy} can keep the set of processed keys in the
state monad (in addition to the \hs{store}), treat processed keys as inputs in
the \hs{getValue} branch of the \hs{fetch} callback, and include the key~\hs{k}
into the set of processed keys in the \hs{putValue} branch. This eliminates
unnecessary work \emph{within a single build}, but the next build needs to
recursively recompute all target's dependencies again even if no inputs changed.
To save work \emph{between builds}, it is necessary to store some build
information~\hs{i} persistently.

\subsection{The Need for Polymorphism in Task}\label{sec-why-polymorphism}

The previous example illustrates why the \hs{Task} abstraction is polymorphic
in~\hs{f}. Recall its definition from~\S\ref{sec-task}:

\begin{minted}[xleftmargin=10pt]{haskell}
newtype Task c k v = Task (@\std{forall}@ f. c f => (k -> f v) -> f v)
\end{minted}

@@

But the task description cannot possibly work for \emph{any} \hs{f} whatsoever; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c} and, in exchange, the task has access to the operations of class \hs{c}.
So the type \hs{Task} emerges naturally, almost inevitably.
But now that it \emph{has} emerged, we find that constraints \hs{c} classify
task descriptions in a very interesting, and practically useful, way:

\begin{itemize}
\item \hs{Task}~\hs{Applicative}: In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}: As we shall see in \S\ref{sec-task-monad}, a monadic
  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
  \cmd{D}. A simple example of a task with dynamic dependencies is \Excel
  formula \cmd{INDIRECT("A" \& C1)} from~\S\ref{sec-background-excel}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: a functorial task
  description cannot even use the application operator \hs{<*>}, which limits
  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}
  (ignoring the recent multi-stage builds). It
  is interesting to note that, when run on such a task description, the
  \hs{busy} build system will build each key at most once, thus partially
  fulfilling the minimality requirement~\ref{def-minimal}. Alas, it still has no
  mechanism to decide which input keys changed since the previous build.
\item \hs{Task}~\hs{Selective} corresponds to task descriptions with
  \emph{conditional statements}, e.g. \Excel formula \cmd{IF(C1=1,B2,A2)},
  where it is possible to statically \emph{over-approximate} the set of task
  dependencies. Here \hs{Selective} is a type class of \emph{selective
  applicative functors}~\cite{mokhov_selective_2019}, which allows us to model
  build systems like \Dune~\cite{dune} using the presented framework.
\item \hs{Task}~\hs{MonadFail} corresponds to monadic tasks that may fail. For
  example, the formula \cmd{A1/A2} may fail due to division by zero. We will
  discuss this in~\S\ref{sec-failures}.
\item \hs{Task}~\hs{MonadPlus}, \hs{Task}~\hs{MonadRandom} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

@@

\noindent
The big difference compared to \hs{sprsh1} is that the computation now takes
place in a \hs{Monad}, which allows us to extract the value of \hs{c1} and
\hs{fetch} \emph{different keys} depending on whether or not
\hs{c1}~\hs{==}~\hs{1}. Note that in this example one can statically determine
the sets of possible dependencies of the formulae \cmd{IF(C1=1,B2,A2)} and
\cmd{IF(C1=1,A1,B1)} but this cannot be done in general -- recall the
spreadsheet with the formula \cmd{INDIRECT("A" \& C1)}
from~\S\ref{sec-background-excel}, where the argument of the \cmd{INDIRECT}
function is a string computed dynamically during the build. Such tasks can also
be captured using \hs{Tasks}~\hs{Monad}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh3 :: Tasks Monad String Integer
sprsh3 "B1" = Just $ Task $ \fetch -> do
    c1 <- fetch "C1"
    fetch ("A" ++ show c1)
sprsh3 _ = Nothing
\end{minted}
\vspace{1mm}

\noindent
Since the \hs{busy} build system introduced in~\S\ref{sec-general-build} always
rebuilds every dependency it encounters, it is easy for it to handle dynamic
dependencies. For minimal build systems, however, dynamic dependencies, and hence
monadic tasks, are much more challenging, as we shall see
in~\S\ref{sec-implementations}.

@@

\noindent
Here we do not need any effects in the \hs{fetch} callback to \hs{task}, so
we can use the standard Haskell \hs{Identity} monad (Fig.~\ref{fig-stdlib}).
This is another use of polymorphism in \hs{f}, discussed in~\S\ref{sec-why-polymorphism}.
The use of \hs{Identity} just fixes the ``impedance mismatch'' between the
function \hs{getValue}, which returns a pure value~\hs{v}, and the \hs{fetch}
argument of the \hs{task}, which must return an \hs{f}~\hs{v} for some \hs{f}.
To fix the mismatch, we wrap the result of \hs{getValue} in the \hs{Identity}
monad and pass to the \hs{task}. The result has type \hs{Identity}~\hs{v}, which
we unwrap with \hs{runIdentity}.

@@

The presented definition of correctness needs to be adjusted for build systems
that support non-deterministic tasks and shallow cloud builds, as will be
discussed in sections~\S\ref{sec-non-determinism} and~\S\ref{sec-cloud-aspects},
respectively.

@@

\noindent
Here \hs{Const} is a standard Haskell type defined in Fig.~\ref{fig-stdlib}. We
instantiate \hs{f} to \hs{Const}~\hs{[@@k]}. So a value of type \hs{f}~\hs{v},
or in this case \hs{Const}~\hs{[@@k]}~\hs{v}, contains no value \hs{v}, but does
contain a list of keys of type \hs{[@@k]} which we use to record dependencies.
The \hs{fetch} callback that we pass to \hs{task} records a single dependency;
and the standard definition of \hs{Applicative} for \hs{Const} (which we give
in Fig.~\ref{fig-stdlib}) combines the dependencies from different parts of the
task. Running the task with \hs{f}~=~\hs{Const}~\hs{[@@k]} will thus
accumulate a list of the task's dependencies -- and that is what
\hs{dependencies} does:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B1"
["A1","A2"]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies $ fromJust $ sprsh1 "B2"
["B1"]
\end{minted}
\vspace{1mm}

@@

Here is an implementation of \hs{track} based on the standard Haskell
\hs{WriterT} \emph{monad transformer}~\cite{liang1995monad}, whose main
types are listed in Fig.~\ref{fig-stdlib}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Monad m => Task Monad k v -> (k -> m v) -> m (v, [(k, v)])
track task fetch = runWriterT $ run task trackingFetch
  where
    trackingFetch :: k -> WriterT [(k, v)] m v
    trackingFetch k = do v <- lift@\,@(fetch k); tell@\,@[(k, v)]; return@\,\blk{v}@
\end{minted}
\vspace{1mm}

\noindent
This function uses the \hs{WriterT} transformer for recording additional
information~--~a list of key/value pairs \hs{[(@@k,}~\hs{v)]}~--~when executing
a task in an arbitrary monad~\hs{m}. We substitute the given \hs{fetch} with a
\hs{trackingFetch} that, in addition to fetching a value, tracks the
corresponding key/value pair. The \hs{task} returns a value of type
\hs{WriterT}~\hs{[(@@k,}~\hs{v)]}~\hs{m}~\hs{v}, which we unwrap with
\hs{runWriterT}. We will use \hs{track} when implementing monadic build systems
with dynamic dependencies, see~\S\ref{sec-implementation-shake}.

Here we show an example of \hs{track}ing monadic tasks when
\hs{m}~\hs{=}~\hs{IO}, by defining a corresponding \hs{fetchIO} of type
\hs{String}~\hs{->}~\hs{IO}~\hs{Integer}, which allows us to demonstrate the
dynamic nature of monadic dependencies in GHCi.

@@

Here the keys $n < 2$ are input parameters, and one can obtain the usual
Fibonacci sequence by picking $F_0=0$ and $F_1=1$, respectively. Any minimal
build system will compute the sequence with memoization, i.e. without
recomputing the same value twice.

@@

\noindent
The main case ($m>0 \wedge n>0$) starts by looking up the pair of indices
\hs{(}\hs{m,}~\hs{n)} in the \hs{cache}. If the cache has \hs{Nothing}, we
calculate the resulting \hs{value} as before and \hs{modify} the cache
accordingly; otherwise, if we have a cache hit, we return the obtained value
immediately, skipping the actual calculation and thus potentially saving a large
amount of work. Indeed you do not want to recompute $A(4,2)=2^{65536}-3$
unnecessarily; all of its 19,729 decimal digits have already been helpfully
computed, e.g. see~Kosara~\shortcite{ackermann42}. We will use \hs{MonadState}
tasks in our models of cloud build systems in~\S\ref{sec-implementations}.

### 2019-10-19 papers/jfp/4-schedulers.tex
Polish on S3-4

Section \S\ref{sec-background} introduced three different \emph{task schedulers}
that decide which tasks to execute and in what order; see the ``Scheduler''
column of Table~\ref{tab-summary} in \S\ref{sec-background-summary}. This
subsection explores the properties of the three schedulers, and possible
implementations.

@@

The topological scheduler pre-computes a linear order of tasks, which when
followed ensures dependencies are satisfied, then executes the required tasks in
that order. Computing such a linear order is straightforward -- given a task
description and output \hs{key} first find the (acyclic) graph of the \hs{key}'s
dependencies, then compute a topological order. Taking the \Make example from
Fig.~\ref{fig-make}, we might compute the order:

\begin{enumerate}
\item \cmd{util.o}
\item \cmd{main.o}
\item \cmd{main.exe}
\end{enumerate}

\noindent
Given the dependencies, we could have equally chosen to build \cmd{main.o}
first, but \cmd{main.exe} \emph{must} come last.

The advantage of this scheme is simplicity -- compute an order, then execute
tasks in that order. In addition any missing nodes or cycles can be detected
from the graph, and reported to the user before any work has commenced.

@@

To handle dynamic dependencies we cannot precompute a static order -- we must
interleave running tasks and ordering tasks. One approach is just to build tasks
in an arbitrary order, but if a task calls \hs{fetch} on an out-of-date key
\cmd{dep}, abort the task and build \cmd{dep} instead. Returning to the example
from Fig.~\ref{fig-make}, we might build in the order:

\begin{enumerate}
\item \cmd{main.exe} (abort because it requires \cmd{util.o})
\item \cmd{util.o}
\item \cmd{main.o}
\item \cmd{main.exe} (restart from scratch, completing successfully this time)
\end{enumerate}

\noindent
We start with \cmd{main.exe} (chosen randomly), but discover it requires
\cmd{util.o}, so instead start building \cmd{util.o}. Next we choose to build
\cmd{main.o} (again, randomly), before finally returning to \cmd{main.exe},
which now has all its dependencies available and completes without error.

The approach works, but has a number of disadvantages. Firstly, it requires a
technical mechanism to abort a task -- easy enough in our theoretical setting of
\hs{Task} (see the implementation in~\S\ref{sec-implementation-excel}) but leads
to engineering concerns in the real world. Secondly, it is not minimal in the
sense that a task may start, do some meaningful work, and then abort, repeating
that same work when restarted.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
they are requested, suspending the currently running task. Using the example
from Fig.~\ref{fig-make}, we would build:

\begin{itemize}
\item \cmd{main.exe} (suspended)\\
  $\hookrightarrow$ \cmd{util.o}
\item \cmd{main.exe} (resumed then suspended again)\\
  $\hookrightarrow$   \cmd{main.o}
\item \cmd{main.exe} (completed)
\end{itemize}

\noindent
As we see, \cmd{main.exe} is started first as it is the required target. It soon
discovers a dependency on \cmd{util.o} and suspends to build \cmd{util.o},
resumes a build before building \cmd{main.o} then continues its work to finish
\cmd{main.exe}.

This scheduler (when combined with a suitable rebuilder) provides a minimal
build system which supports dynamic dependencies. In our model, a suspending
scheduler is easy to write -- it makes a function call to compute each
dependency. However, a more practical implementation is likely to build multiple
dependencies in parallel, which then requires a more explicit task suspension
and resumption. To implement suspension there are two standard approaches:

\begin{itemize}
\item Blocking threads or processes. This approach is relatively easy, but can
require significant resources, especially if a large number of tasks are
suspended. In languages with cheap green threads (e.g. Haskell) the approach is
more feasible, and it was the original approach taken by \Shake.
\item Continuation-passing style~\cite{claessen_continuations} can allow the
remainder of a task to be captured, paused, and resumed later. Continuation
passing is efficient, but requires the build script to be architected to allow
capturing continuations. \Shake currently uses continuations.
\end{itemize}

\noindent
While a suspending scheduler is theoretically optimal, it is only better in
practice than a restarting scheduler if the cost of avoided duplicate work
outweighs the cost of suspending tasks. Note furthermore that the cost of
duplicate work may often be just a fraction of the overall build cost.

>>>>

Section \S\ref{sec-background} introduced three different \emph{task schedulers}
that decide which tasks to execute and in what order; see the ``Scheduler''
column of Table~\ref{tab-summary} in \S\ref{sec-background-summary}. The
following subsections explore the properties of the three schedulers, and
possible implementations.

@@

The topological scheduler pre-computes a linear order of tasks, which when
followed ensures dependencies are satisfied, then executes the required tasks in
that order. Computing such a linear order is straightforward -- given a task
description and a target \hs{key}, first find the (acyclic) graph of the
\hs{key}'s dependencies, then compute a topological order. Taking the \Make
example from Fig.~\ref{fig-make}, we might compute the order:

\begin{enumerate}
\item \cmd{main.o}
\item \cmd{util.o}
\item \cmd{main.exe}
\end{enumerate}

\noindent
Given the dependencies, we could have equally chosen to build \cmd{util.o}
first, but \cmd{main.exe} \emph{must} come last.

The advantage of this scheme is simplicity -- compute an order, then execute
tasks in that order. In addition, any missing keys or dependency cycles can be
detected from the graph, and reported to the user before any work has commenced.

@@

To handle dynamic dependencies we cannot precompute a static order -- we must
interleave running tasks and ordering tasks. One approach is just to build tasks
in an arbitrary order, and if a task calls \hs{fetch} on an out-of-date key
\cmd{dep}, abort the task and build \cmd{dep} instead. Returning to the example
from Fig.~\ref{fig-make}, we might build the tasks in the order:

\begin{enumerate}
\item \cmd{main.exe} (abort because it depends on \cmd{util.o} which is out of
      date)
\item \cmd{main.o}
\item \cmd{util.o}
\item \cmd{main.exe} (restart from scratch, completing successfully this time)
\end{enumerate}

\noindent
We start with \cmd{main.exe} (chosen randomly), but discover it depends on
\cmd{main.o}, so instead start building \cmd{main.o}. Next we choose to build
\cmd{util.o} (again, randomly), before finally returning to \cmd{main.exe} that
now has all its dependencies available and completes successfully.

The approach works, but has a number of disadvantages. Firstly, it requires a
technical mechanism to abort a task, which is easy in our theoretical setting
with \hs{Task} (see an implementation in~\S\ref{sec-implementation-excel}) but
leads to engineering concerns in the real world. Secondly, it is not minimal in
the sense that a task may start, do some meaningful work, and then abort,
repeating that same work when restarted.

@@

An alternative approach, utilised by the \hs{busy} build system
(\S\ref{sec-general-build}) and \Shake, is to simply build dependencies when
they are requested, suspending the currently running task when needed. Using the
example from Fig.~\ref{fig-make}, we would build:

\begin{itemize}
\item \cmd{main.exe} (suspended)\\
  $\hookrightarrow$ \cmd{main.o}
\item \cmd{main.exe} (resumed then suspended again)\\
  $\hookrightarrow$   \cmd{util.o}
\item \cmd{main.exe} (completed)
\end{itemize}

\noindent
We start building \cmd{main.exe} first as it is the required target. We soon
discover a dependency on \cmd{main.o} and suspend the current task
\cmd{main.exe} to build \cmd{main.o}, then resume and suspend again to build
\cmd{util.o}, and finally complete the target \cmd{main.exe}.

This scheduler (when combined with a suitable rebuilder) provides a minimal
build system that supports dynamic dependencies. In our model, a suspending
scheduler is easy to write -- it makes a function call to compute each
dependency. However, a more practical implementation is likely to build multiple
dependencies in parallel, which then requires a more explicit task suspension
and resumption. To implement suspension there are two standard approaches:

\begin{itemize}
\item Blocking threads or processes. This approach is relatively easy, but can
require significant resources, especially if a large number of tasks are
suspended. In languages with cheap green threads (e.g. Haskell) the approach is
more feasible, and it was the original approach taken by \Shake.
\item Continuation-passing style~\cite{claessen_continuations} can allow the
remainder of a task to be captured, paused, and resumed later. Continuation
passing is efficient, but requires the build script to be architected to allow
capturing continuations. \Shake currently uses this approach.
\end{itemize}

\noindent
While a suspending scheduler is theoretically optimal, in practice it is better
than a restarting scheduler only if the cost of avoided duplicate work
outweighs the cost of suspending tasks. Note furthermore that the cost of
duplicate work may often be just a fraction of the overall build cost.

### 2019-10-19 papers/jfp/8-engineering.tex
Polish on S3-4

This approach is convenient if \emph{tasks are not aware of failures}, e.g. we
can model \Excel formulas as pure arithmetic functions, and introduce failures
``for free'' if/when needed by instantiating \hs{Tasks} with an appropriate
\hs{f}. In a real system this \hs{f} would be more complex than just \hs{Maybe},
for example \hs{MaybeT}~\hs{(State}~\hs{Spreadsheet)}, thus allowing us to
combine failures with access to \Excel's spreadsheet state by using the
\hs{MaybeT} monad transformer~\cite{liang1995monad}.

@@

sprsh3 :: Tasks MonadFail String Integer
sprsh3 "B1" = Just $ Task $ \fetch -> do
    a1 <- fetch "A1"
    a2 <- fetch "A2"
    if a2 == 0 then fail "division by 0" else return (a1 `div` a2)
sprsh3 _ = Nothing
\end{minted}
\vspace{1mm}

@@

\begin{minted}[xleftmargin=10pt]{haskell}
sprsh4 :: Tasks MonadRandom String Integer
sprsh4 "B1" = Just $ Task $ \fetch -> do a1 <- fetch "A1"
                                         r  <- getRandom (1,2)
                                         return (a1 + r)
sprsh4 _    = Nothing
\end{minted}
\vspace{1mm}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh5 "B1" = Just $ Task $ \fetch -> do
    formula <- fetch "B1-formula"
    evalFormula fetch formula
\end{minted}
\vspace{1mm}

@@

\noindent
The function \hs{selfTracking} defined below is a generalisation of the approach
explained in the above \Excel example \hs{sprsh5}. The function takes a parser
for scripts, of type \hs{s}~\hs{->}~\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}, and a
description of \emph{how to build all scripts}, of type
\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{s}. For \hs{sprsh5}, the latter would simply
fetch \cmd{B1-formula} when given the key \cmd{B1} and return \hs{Nothing}
otherwise, but the presented approach can cope with much more sophisticated
scenarios where scripts themselves are derived from ``script sources'', e.g. all
C compilation scripts can be obtained from a single pattern rule, such as
\cmd{gcc -c [source] -o [object]}. The resulting typed task description
\hs{TasksT}~\hs{Monad}~\hs{(Key}~\hs{k}~\hs{v}~\hs{s)} tracks both values and
scripts that compute them, and is therefore self-tracking.

>>>>

This approach is convenient if \emph{tasks are not aware of failures}, e.g. we
can model \Excel formulae as pure arithmetic functions, and introduce failures
``for free'' if/when needed by instantiating \hs{Tasks} with an appropriate
\hs{f}. In a real system this \hs{f} would be more complex than just \hs{Maybe},
for example \hs{MaybeT}~\hs{(State}~\hs{Spreadsheet)}, thus allowing us to
combine failures with access to \Excel's spreadsheet state by using the
\hs{MaybeT} monad transformer~\cite{liang1995monad}.

@@

sprsh4 :: Tasks MonadFail String Integer
sprsh4 "B1" = Just $ Task $ \fetch -> do
    a1 <- fetch "A1"
    a2 <- fetch "A2"
    if a2 == 0 then fail "division by 0" else return (a1 `div` a2)
sprsh4 _ = Nothing
\end{minted}
\vspace{1mm}

@@

\begin{minted}[xleftmargin=10pt]{haskell}
sprsh5 :: Tasks MonadRandom String Integer
sprsh5 "B1" = Just $ Task $ \fetch -> do a1 <- fetch "A1"
                                         r  <- getRandom (1,2)
                                         return (a1 + r)
sprsh5 _    = Nothing
\end{minted}
\vspace{1mm}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh6 "B1" = Just $ Task $ \fetch -> do
    formula <- fetch "B1-formula"
    evalFormula fetch formula
\end{minted}
\vspace{1mm}

@@

\noindent
The function \hs{selfTracking} defined below is a generalisation of the approach
explained in the above \Excel example \hs{sprsh6}. The function takes a parser
for scripts, of type \hs{s}~\hs{->}~\hs{Task}~\hs{Monad}~\hs{k}~\hs{v}, and a
description of \emph{how to build all scripts}, of type
\hs{Tasks}~\hs{Monad}~\hs{k}~\hs{s}. For \hs{sprsh6}, the latter would simply
fetch \cmd{B1-formula} when given the key \cmd{B1} and return \hs{Nothing}
otherwise, but the presented approach can cope with much more sophisticated
scenarios where scripts themselves are derived from ``script sources'', e.g. all
C compilation scripts can be obtained from a single pattern rule, such as
\cmd{gcc -c [source] -o [object]}. The resulting typed task description
\hs{TasksT}~\hs{Monad}~\hs{(Key}~\hs{k}~\hs{v}~\hs{s)} tracks both values and
scripts that compute them, and is therefore self-tracking.
